---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Since repository itself provide zip file, we will not download it from Internet.
If you don't have [data file](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) for some reason, you can download it.

Also, make note -- we will need plyr and ggplot2 packages, so make sure they are present. If not -- you should install them using ```install.packages```

Since data file is zipped -- we unzip it using `unz` function before read.
```{r}
data <- read.csv(unz("activity.zip", "activity.csv"))
head(data)
```

Now lets check column classes and fix them if need
```{r}
str(data)
data$date <- as.Date(data$date, format="%Y-%m-%d")
```

Last step: since I'm russian, lets set locale to english to not fear you with cyrillic
```{r}
Sys.setlocale("LC_ALL","English")
```

Ok, for now dataset is ready for manipulations.

## What is mean total number of steps taken per day?
For calculatinf this we summarize steps done within same date and then calculate mean value.

```{r}
steps_per_day <- aggregate(steps ~ date, data=data, FUN=sum)
```
Here is total steps per day frequence diagram.
```{r, echo= FALSE}
hist(steps_per_day$steps, xlab = 'steps per day', main = "total steps per day frequency", breaks=20)
```

Now we're ready to calculate mean and median values:
```{r}
mean(steps_per_day$steps, na.rm = TRUE)
median(steps_per_day$steps)
```

## What is the average daily activity pattern?

To answer this question first we need to calculate average values for each interval
```{r}
steps_per_interval <- aggregate(steps ~ interval, data=data, FUN=mean)
```

Now lets plot the diagram.
```{r}
plot(steps_per_interval$interval, steps_per_interval$steps, type='l', ylab="average steps count", xlab='interval number', main='typical pattern across all days')
```

So the pattern looks like this: 

* almost no activity till 500th minute
* then steady increasing till 800th minute
* then rapid increase and immediate fall in interval 800 - 1000 min
* 'saw'-style activity till approx. 1900th minute
* then steady decrease

In general, looks like most steps are taken approximately at 850th-900th minute.

## Imputing missing values

Lets do quick analyze for missing values. First, check, how many NA values we have:

```{r}
missing <- sum(is.na(data$steps))
present <- sum(!is.na(data$steps))
percentage <- missing * 100 / (missing + present)
print(missing)
print(percentage)
```
So, 13% of values is missing. Quite big value, lets emulate them, using median for this time interval across all days
(credantisal for this method goes to [stackoverflow auther](https://stackoverflow.com/questions/9322773/how-to-replace-na-with-mean-by-subset-in-r-impute-with-plyr))

```{r}
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
data2 <- ddply(data, ~ interval, transform, steps = impute.mean(steps))
head(data2)
```

Now rebuild frequency histogram and recalculate mean / median values:
```{r}
steps_per_day <- aggregate(steps ~ date, data=data2, FUN=sum)
mean(steps_per_day$steps, na.rm = TRUE)
median(steps_per_day$steps)
hist(steps_per_day$steps, xlab = 'steps per day', main = "total steps per day frequency", breaks=20)
```

So, while mean and median are left same (probably because NAs replaced with means), total average a bit different.
Our 'fix' moved total probability toward more euiality, lowering most probable scenario and increasing less probable.
(Err, sorry, english wasn't my favorite classes in school.)

## Are there differences in activity patterns between weekdays and weekends?

To answer last questin, first  add new column to our dataset, containing weekday name:
```{r}
data2$wd <- weekdays(data2$date, abbreviate = T)
data2$weekend <-  factor(data2$wd %in% c("Sun", "Sat"), levels=c(FALSE, TRUE), labels=c("weekday", "weekend"))
head(data2)
```

Now calculate mean values for steps per each interval and plot the result,  separately for weekend and weekday:
```{r}
library(ggplot2)
steps_per_interval <- aggregate(steps ~ interval + weekend, data=data2, FUN=mean)
qplot(interval, steps, data=steps_per_interval, facets = weekend ~ ., col=weekend, geom = c("point", "path"))
```